Stack Safety is a Security Property

### 2021-04-15 tex/main.tex
example 3 formatting
Moved passing examples to after some formalism

\paragraph*{Pass by Reference and Stack Arguments}
\bcp{I wonder if this could be postponed so that we get to the formal
  definitions for the basic property sooner...}

\begin{figure}
  \centering
  \begin{subfigure}{.3\textwidth}
\begin{verbatim}
int main() {
0:           int x;
1:           x = 5;
2[P($SP)C]:  x = f(x);
3[R]:        return x;
}

int f(int a) {
10:    int y;
11:    y = 2*a;
12:    a = 0;
13[R]: return y;
}
\end{verbatim}
  \end{subfigure}
\begin{subfigure}{.65\textwidth}
\begin{center}
\begin{tabular}{l r | l}
  {\tt 2} &
  \multicolumn{2}{c}{
    \memoryaddrs[(0)]{8em}
    \memory{4}{\unsealc}
    ~$\cdots$
    \MemoryLabel{-19em}{0.75em}{5}
    \MemoryLabel{-14em}{0.75em}{0}
    \MemoryLabel{-10em}{0.75em}{0}
    \MemoryLabel{-7em}{0.75em}{0}
    \vspace{.5em}
  } \\
  {\tt 11} &
  \memoryaddrs[(1)]{20em}
  \memory{1}{\mainsealc}[Seal(0)]%
  \memory{1}{\mainpassc}[P(0-1)]%
  \memory{1}{\retptrc}[RetPtr]%
  \memory{1}{\unsealc}
  ~$\cdots$
  \MemoryLabel{-19em}{0.75em}{5}
  \MemoryLabel{-14em}{0.75em}{5}
  \MemoryLabel{-6em}{0.75em}{0}
  &
  \memoryaddrs[(1)]{20em}
  \memory{1}{\mainsealc}
  \memory{1}{\mainpassc}
  \memory{1}{\retptrc}
  \memory{1}{\unsealc}
  ~$\cdots$
  \MemoryLabel{-19em}{0.75em}{\(v_0\)}
  \MemoryLabel{-14em}{0.75em}{5}
  \MemoryLabel{-6em}{0.75em}{\(v_1\)}
  \\
  {\tt 12} &
  \memoryaddrs[(1)]{20em}
  \memory{1}{\mainsealc}
  \memory{1}{\mainpassc}
  \memory{1}{\retptrc}
  \memory{1}{\unsealc}
  ~$\cdots$
  \MemoryLabel{-19em}{0.75em}{5}
  \MemoryLabel{-14em}{0.75em}{5}
  \MemoryLabel{-6em}{0.75em}{10}
  &
  \memoryaddrs[(1)]{20em}
  \memory{1}{\mainsealc}
  \memory{1}{\mainpassc}
  \memory{1}{\retptrc}
  \memory{1}{\unsealc}
  ~$\cdots$
  \MemoryLabel{-19em}{0.75em}{\(v_0\)}
  \MemoryLabel{-14em}{0.75em}{5}
  \MemoryLabel{-6em}{0.75em}{10}
  \\
  {\tt 13} &
  \memoryaddrs[(1)]{20em}
  \memory{1}{\mainsealc}
  \memory{1}{\mainpassc}
  \memory{1}{\retptrc}
  \memory{1}{\unsealc}
  ~$\cdots$
  \MemoryLabel{-19em}{0.75em}{5}
  \MemoryLabel{-14em}{0.75em}{0}
  \MemoryLabel{-6em}{0.75em}{10}
  &
  \memoryaddrs[(1)]{20em}
  \memory{1}{\mainsealc}
  \memory{1}{\mainpassc}
  \memory{1}{\retptrc}
  \memory{1}{\unsealc}
  ~$\cdots$
  \MemoryLabel{-19em}{0.75em}{\(v_0\)}
  \MemoryLabel{-14em}{0.75em}{0}
  \MemoryLabel{-6em}{0.75em}{10}
  \\
  {\tt 3} &
  \multicolumn{2}{c}{
    \memoryaddrs[(0)]{8em}
    \memory{4}{\unsealc}
    ~$\cdots$
    \MemoryLabel{-19em}{0.75em}{10}
    \MemoryLabel{-14em}{0.75em}{0}
    \MemoryLabel{-10em}{0.75em}{3}
    \MemoryLabel{-7em}{0.75em}{10}
    \vspace{.5em}
  } \\
\end{tabular}
\end{center}

int main() {
0:         int x, y;
1[P(x)C]:  f(x);
2:         x = 0;
3[P(y)C]:  f(y);
4[R]:      return x;
}

void f(int &a) {
10:    if(!stash)
11:      stash = a;
12:    *stash = 5;
13[R]: return;
}
\end{verbatim}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
    \begin{tabular}{l l}
      {\tt 10} &
      \memoryaddrs[(0)]{17em}
      \memory{1}{\mainpassc}[{\makebox[0pt]{P(0-1)}}]%
      \memory{1}{\mainsealc}[{\makebox[0pt]{Seal(0)}}]%
      \memory{1}{\retptrc}
      \memory{1}{\unsealc}
      ~$\cdots$ \\
      {\tt 13} &
      \memoryaddrs[(1)]{17em}
      \memory{1}{\mainpassc}
      \memory{1}{\mainsealc}
      \memory{1}{\retptrc}
      \memory{1}{\unsealc}
      ~$\cdots$
      \MemoryLabel{-18em}{0.75em}{5}
      \\
      {\tt 3} &
      \memoryaddrs[(0)]{12em}
      \memory{4}{\unsealc}
      ~$\cdots$
      \MemoryLabel{-18em}{0.75em}{0}
      \\
      {\tt 10} &
      \memoryaddrs[(0)]{17em}
      \memory{1}{\mainsealc}[{\makebox[0pt]{Seal(0)}}]%
      \memory{1}{\mainpassc}[{\makebox[0pt]{P(0-1)}}]%
      \memory{1}{\retptrc}
      \memory{1}{\unsealc}
      ~$\cdots$
      \MemoryLabel{-18em}{0.75em}{0}
      \\
      {\tt 13} &
      \memoryaddrs[(0)]{17em}
      \memory{1}{\mainsealc}
      \memory{1}{\mainpassc}
      \memory{1}{\retptrc}
      \memory{1}{\unsealc}
      ~$\cdots$
      \MemoryLabel{-18em}{0.75em}{\bf 5}
\end{tabular}
\end{subfigure}
\caption{A violation with pass-by-reference}
\label{fig:passing}
\end{figure}

To this point we have ignored function arguments, since our simplest model places them in
registers. This is an oversimplification, which we will now correct, starting with the simple
case: passing by value on the stack. Example \ref{fig:passingsimple} shows a sample trace of
a program that passes a value on the stack, without an attack.
At the call, we add a new annotation \(P(\$SP)\) identifying a memory location to be
passed, and our diagram displays it in magenta. The caller's frame locations that are not
passed are sealed as usual. Just as we marked sealed memory with the depth of its owner\bcp{??},
we do the same for passed memory -- but it has multiple owners. We keep track of the
current depth of the active function next to the stack pointer in the diagram, for reference.
The argument {\tt a} has
its value copied from {\tt x}, and we note that {\tt f} can read it -- because it is marked
as passed between depths 0 and 1, it is not varied. Then from line 12 to 13 it is even
modified, and naturally this is permitted as well.

Our model provides only limited support for handling fully first-class pointers into
stack frames, e.g., pointers obtained by taking the address of a local variable.
Unlike the case for pass-by-reference, there is no easy way to model the intended sharing behavior
of such pointers in terms of stack frame activation patterns.  The best approximation we can make is to
treat an address-taken variable in the same way as a passed argument, with the additional property
that it remains accessible at \emph{all} stack depths above the caller.
This prevents most violations of temporal safety, although stale pointers can still be
used in certain corner cases. \apt{Trying not to spell out the details, which are complex.}
More seriously, callees are permitted to access the location without having
been explicitly given a pointer to it.  Preventing the latter kind of abuse
is the provence of a fine-grained memory safety policy, e.g. based on capabilities
(implemented by tags or otherwise); we argue it is beyond the scope of stack safety
\emph{per se}.

@@

We are finally ready for our definition of stack safety. We begin
by describing how we model knowledge about the program structure,
then dive into the formal definitions of local state encapsulation and
well-bracketed control flow, which together constitute stack safety\bcp{not
  any more?}.
Initially we will focus on a simple stack model, with a single stack
that doesn't share between caller and callee. All arguments are passed
in registers.

We begin with our only annotations being {\em call} or {\em return}, and
define our code map accordingly.

\[\ann \in \ANNS ::= \callmap + \retmap\]
\[\codemap \in \CODEMAPS ::= \WORDS \rightharpoonup \ANNS\]

When a function makes a call, it establishes that its frame should be
``sealed'' for later use. At any given step of a program, any number of
activations may have claimed memory in this way, but which cells are sealed
is dynamic information and cannot be deduced from the state alone. So it must
be tracked separately. We must also track the conditions under which a caller
considers itself to have been returned to.

Each system component will be assigned to a {\em domain}, which may be {\em Outside}
of the stack entirely, {\em Unsealed} and therefore accessible, or {\em Sealed}
at a particular depth\bcp{What about code??}. Each caller will register a {\em target} condition -- a
predicate on machine states that must hold for it to consider itself active.
A context is a pair of a ``domain map'' from components to domains, and a
``return target list,'' which is a stack of targets, one for each caller awaiting
return.  \ifaftersubmission\bcp{spacing is funny}\fi

@@

We parameterize over the structure of the stack, particularly the means by
which a caller identifies the data that should be protected during a call. We term
this the {\em sealing convention}, \(\sealcon \subseteq \MACHS \times \COMPONENTS\), which
maps a machine state to a set of addresses.
If a state's \(\PCname\) is annotated as a call, the sealing convention tells us
which addresses must be protected until the return. Similarly, we abstract over what
it means to return with a {\em return convention}, \(\retcon \subseteq \MACHS \times \MACHS\),
which relates

Now we can define the context update relation. Let \(\sealcon\) be the sealing convention
that maps a machine state to a set of components, typically the set of addresses below
the stack pointer. Similarly, \(\retcon ~ \mach ~ \mach'\) will be the return convention
that relates the call state \(\mach\) to return states \(\mach'\) that are valid returns.
The usual return convention is that \(\mach\) and \(\mach'\) agree on the stack pointer,
and \(\mach'\) has its program counter advanced by one instruction from \(\mach\).
Then at each step we update the context according to the context-step function,
\(\stepsto_C\).

The rules for \(\stepsto_C\) are given inductively, and apply in the listed order.
A call simply pushes the return target onto the return stack and seals all components
with the depth of the previous return stack. A return
recursively searches for a valid return target at the next step, and unseals
every component sealed at a depth that it passes.

@@

\judgmenttwobr[ReturnFound]
              {\(\codemap ~ (\mach ~ \PCname) = \retmap\)}
              {\(\mach \stepsto \mach' \in \target\)}
              {\(\components = \{\component | \domMap ~ \component = \sealed{|\rts|}\}\)}
              {\(\domMap' = \domMap[\components \mapsto \unsealed]\)}
              {\(\mach,(\domMap,\target::\rts) \stepsto_C \domMap',\rts\)}

@@

>>>>

@@

We are finally ready for our definition of stack safety. We begin
by describing how we model knowledge about the program structure,
then dive into the formal definitions of local state encapsulation and
well-bracketed control flow, which together constitute stack safety\bcp{not
  any more?}.
Initially we will focus on a simple stack model, with a single stack
that doesn't share between caller and callee. All arguments are passed
in registers. Recall that we keep track of which code addresses are
annotated as calls, with a ``code map'':
\[\codemap \in \CODEMAPS \subseteq \WORDS\]
When a function makes a call, it establishes that its frame should be
``sealed'' for later use. At any given step of a program, any number of
activations may have claimed memory in this way, but which cells are sealed
is dynamic information and cannot be deduced from the state alone. So it must
be tracked separately. We must also track the conditions under which a caller
considers itself to have been returned to.

Each system component will be assigned to a {\em domain}, which may be {\em Outside}
of the stack entirely, {\em Unsealed} and therefore accessible, or {\em Sealed}
at a particular depth. Each caller will register a {\em target}
condition -- a predicate on machine states that must hold for it to consider itself active.
A context is a pair of a ``domain map'' from components to domains, and a
``return target list,'' which is a stack of targets, one for each caller awaiting
return.  \ifaftersubmission\bcp{spacing is funny}\fi

@@

We parameterize over the structure of the stack, particularly the means by
which a caller identifies the data that should be protected during a call. We term
this the {\em sealing convention}, \(\sealcon \subseteq \MACHS \times \COMPONENTS\), which
maps a machine state to a set of addresses.
If a state's \(\PCname\) is annotated as a call, the sealing convention tells us
which addresses must be protected until the return. Similarly, we abstract over what
it means to return with a {\em return convention}, \(\retcon \subseteq \MACHS \times \MACHS\),
which relates call states to the states that can be considered valid returns to their caller.
The usual return convention is that \(\mach\) and \(\mach'\) agree on the stack pointer,
and \(\mach'\) has its program counter advanced by one instruction from \(\mach\).
But whatever the return convention, at each step we update the context according
to the context-step function, \(\stepsto_C\).

The rules for \(\stepsto_C\) are given inductively, and apply in the listed order.
A call annotation tells us to push the return target onto the return stack and
seal all components with the depth of the previous return stack. If we aren't making a call,
we identify a return by looking ahead and checking if we're about to step to a state
that matches the topmost return target. If neither is true, the context remains unchanged.

@@

\judgmentthree[ReturnFound]
              {\(\mach \stepsto \mach' \in \target\)}
              {\(\components = \{\component | \domMap ~ \component = \sealed{|\rts|}\}\)}
              {\(\domMap' = \domMap[\components \mapsto \unsealed]\)}
              {\(\mach,(\domMap,\target::\rts) \stepsto_C \domMap',\rts\)}

@@

\begin{figure}
  \centering
  \begin{subfigure}{.3\textwidth}
\begin{verbatim}
int main() {
0:          int x;
1:          x = 5;
2[C($SP)]:  x = f(x);
3[R]:       return x;
}

int f(int a) {
10:    int y;
11:    y = 2*a;
12:    a = 0;
13[R]: return y;
}
\end{verbatim}
  \end{subfigure}
\begin{subfigure}{.65\textwidth}
\begin{center}
\begin{tabular}{l r | l}
  {\tt 2} &
  \multicolumn{2}{c}{
    \memoryaddrs[(0)]{8em}
    \memory{4}{\unsealc}
    ~$\cdots$
    \MemoryLabel{-19em}{0.75em}{5}
    \MemoryLabel{-14em}{0.75em}{0}
    \MemoryLabel{-10em}{0.75em}{0}
    \MemoryLabel{-7em}{0.75em}{0}
    \vspace{.5em}
  } \\
  {\tt 11} &
  \memoryaddrs[(1)]{20em}
  \memory{1}{\mainsealc}[Seal(0)]%
  \memory{1}{\mainpassc}[P(0,1)]%
  \memory{1}{\retptrc}[RetPtr]%
  \memory{1}{\unsealc}
  ~$\cdots$
  \MemoryLabel{-19em}{0.75em}{5}
  \MemoryLabel{-14em}{0.75em}{5}
  \MemoryLabel{-6em}{0.75em}{0}
  &
  \memoryaddrs[(1)]{20em}
  \memory{1}{\mainsealc}
  \memory{1}{\mainpassc}
  \memory{1}{\retptrc}
  \memory{1}{\unsealc}
  ~$\cdots$
  \MemoryLabel{-19em}{0.75em}{\(v_0\)}
  \MemoryLabel{-14em}{0.75em}{5}
  \MemoryLabel{-6em}{0.75em}{\(v_1\)}
  \\
  {\tt 12} &
  \memoryaddrs[(1)]{20em}
  \memory{1}{\mainsealc}
  \memory{1}{\mainpassc}
  \memory{1}{\retptrc}
  \memory{1}{\unsealc}
  ~$\cdots$
  \MemoryLabel{-19em}{0.75em}{5}
  \MemoryLabel{-14em}{0.75em}{5}
  \MemoryLabel{-6em}{0.75em}{10}
  &
  \memoryaddrs[(1)]{20em}
  \memory{1}{\mainsealc}
  \memory{1}{\mainpassc}
  \memory{1}{\retptrc}
  \memory{1}{\unsealc}
  ~$\cdots$
  \MemoryLabel{-19em}{0.75em}{\(v_0\)}
  \MemoryLabel{-14em}{0.75em}{5}
  \MemoryLabel{-6em}{0.75em}{10}
  \\
  {\tt 13} &
  \memoryaddrs[(1)]{20em}
  \memory{1}{\mainsealc}
  \memory{1}{\mainpassc}
  \memory{1}{\retptrc}
  \memory{1}{\unsealc}
  ~$\cdots$
  \MemoryLabel{-19em}{0.75em}{5}
  \MemoryLabel{-14em}{0.75em}{0}
  \MemoryLabel{-6em}{0.75em}{10}
  &
  \memoryaddrs[(1)]{20em}
  \memory{1}{\mainsealc}
  \memory{1}{\mainpassc}
  \memory{1}{\retptrc}
  \memory{1}{\unsealc}
  ~$\cdots$
  \MemoryLabel{-19em}{0.75em}{\(v_0\)}
  \MemoryLabel{-14em}{0.75em}{0}
  \MemoryLabel{-6em}{0.75em}{10}
  \\
  {\tt 3} &
  \multicolumn{2}{c}{
    \memoryaddrs[(0)]{8em}
    \memory{4}{\unsealc}
    ~$\cdots$
    \MemoryLabel{-19em}{0.75em}{10}
    \MemoryLabel{-14em}{0.75em}{0}
    \MemoryLabel{-10em}{0.75em}{3}
    \MemoryLabel{-7em}{0.75em}{10}
    \vspace{.5em}
  } \\
\end{tabular}
\end{center}

int main() {
0:         int x, y;
1[C(x)]:  f(x);
2:         x = 0;
3[C(y)]:  f(y);
4[R]:      return x;
}

void f(int &a) {
10:    if(!stash)
11:      stash = a;
12:    *stash = 5;
13[R]: return;
}
\end{verbatim}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
    \begin{tabular}{l l}
      {\tt 10} &
      \memoryaddrs[(0)]{17em}
      \memory{1}{\mainpassc}[{\makebox[0pt]{P(0,1)}}]%
      \memory{1}{\mainsealc}[{\makebox[0pt]{Seal(0)}}]%
      \memory{1}{\retptrc}
      \memory{1}{\unsealc}
      ~$\cdots$ \\
      {\tt 13} &
      \memoryaddrs[(1)]{17em}
      \memory{1}{\mainpassc}
      \memory{1}{\mainsealc}
      \memory{1}{\retptrc}
      \memory{1}{\unsealc}
      ~$\cdots$
      \MemoryLabel{-18em}{0.75em}{5}
      \\
      {\tt 3} &
      \memoryaddrs[(0)]{12em}
      \memory{4}{\unsealc}
      ~$\cdots$
      \MemoryLabel{-18em}{0.75em}{0}
      \\
      {\tt 10} &
      \memoryaddrs[(0)]{17em}
      \memory{1}{\mainsealc}[{\makebox[0pt]{Seal(0)}}]%
      \memory{1}{\mainpassc}[{\makebox[0pt]{P(0-1)}}]%
      \memory{1}{\retptrc}
      \memory{1}{\unsealc}
      ~$\cdots$
      \MemoryLabel{-18em}{0.75em}{0}
      \\
      {\tt 13} &
      \memoryaddrs[(0)]{17em}
      \memory{1}{\mainsealc}
      \memory{1}{\mainpassc}
      \memory{1}{\retptrc}
      \memory{1}{\unsealc}
      ~$\cdots$
      \MemoryLabel{-18em}{0.75em}{\bf 5}
\end{tabular}
\end{subfigure}
\caption{A violation with pass-by-reference}
\label{fig:passing}
\end{figure}

To this point we have ignored function arguments, since our simplest model places them in
registers. This is an oversimplification, which we will now correct, aiming at two kinds
of sharing: passing by value on the stack, and passing by reference. First we will explore
some examples at the same level of detail as our earlier ones, and then move on to
extending our formalization.

Example \ref{fig:passingsimple} shows a sample trace of
a program that passes a value on the stack, without an attack.
We extend our call annotation with an additional parameter identifying a memory location to be
passed, and our diagram displays it in magenta. Our notion of domains is extended as well,
to include \(\passed{\depth,\depth'}\), recording two depths. We note the
current depth of the active function next to the stack pointer in the diagram, for reference.
The argument {\tt a} has
its value copied from {\tt x}, and we note that {\tt f} can read it -- because it is marked
as passed between depths 0 and 1, it is not varied. Then from line 12 to 13 it is even
modified, and naturally this is permitted as well.

Our model provides only limited support for handling fully first-class pointers into
stack frames, e.g., pointers obtained by taking the address of a local variable.
Unlike the case for pass-by-reference, there is no easy way to model the intended
sharing behavior of such pointers in terms of stack frame activation patterns.
The best approximation we can make is to treat an address-taken variable in the
same way as a passed argument, with the additional property
that it remains accessible at \emph{all} stack depths above the caller.
This prevents most violations of temporal safety, although stale pointers can still be
used in certain corner cases.
More seriously, callees are permitted to access the location without having
been explicitly given a pointer to it.  Preventing the latter kind of abuse
is the provence of a fine-grained memory safety policy, e.g. based on capabilities
(implemented by tags or otherwise); we argue it is beyond the scope of stack safety
\emph{per se}.
